% Write your acknowledgements here

%TC:ignore
Phew... what a ride. I don't know how many years these past six years have taken off my lifespan, but what a journey it has been. I don't think I've ever been this busy, exhausted, and frankly at the end of my energy reserves—yet I wouldn't trade this experience for anything.
I still remember that first day in Woodward Buildings, being dropped off by my parents, nervous and uncertain, gradually meeting all the people who would later become my dearest friends. That anxious teenager could never have imagined the person I would become or the incredible minds I would have the privilege to learn alongside.

To Imperial College London, I am deeply grateful for the rigorous teaching that has fundamentally transformed how I see the world. The education I received here didn't just fill my mind with knowledge—it rewired how I think. I love the way I was taught to see connections where others see chaos, to find the hidden beauty in the Fourier Transform, to appreciate the philosophical depths of reinforcement learning.

This is the perfect moment to give thanks to the extraordinary people who shaped this journey:

To Gloria Sun, whose career-oriented mindset and relentless push for excellence ignited an ambition in me I didn't know existed. Your influence helped me realize that settling for "good enough" was never an option—thank you for believing I could achieve more than I believed possible.
To my incredible friends Nabeel and Francisco, who turned the most stressful moments into adventures filled with laughter. Through late-night study sessions, impossible deadlines, and moments when I questioned everything, you were there with encouragement, terrible jokes, and the kind of friendship that makes even the hardest challenges bearable. You reminded me that brilliance is better when shared.

To my girlfriend Eda, whose love has been my anchor through every storm. When the weight of research felt overwhelming, when imposter syndrome crept in, when I forgot to eat or sleep because I was lost in equations—you were there. Your patience, understanding, and unwavering belief in me gave me strength when I had none left. You celebrated every small victory and helped me through every setback. This thesis exists not just because of my work, but because of your love.

To my supervisor, Dr. Anil Bharath, for accepting this ambitious project and giving me the intellectual freedom to explore. Your leniency wasn't just kindness—it was trust in my ability to find my own path through complex problems. Thank you for believing in this research when it was just a wild idea, and for providing the guidance that helped transform that idea into something meaningful.

To my parents, who drove me to that first day at Imperial and have supported every dream since. Your sacrifices, encouragement, and unconditional love laid the foundation for everything I've achieved.

Finally, to the reader of this thesis—thank you for taking interest in this work. 
%TC:endignore